### 2016-06-28 publications/Journal/4a-allocation.tex
Deduplicate the section title
Restructure and write about hotswap component with statistics gathering

\section{Model Allocation}
%TODO explain model allocation
%TODO why model allocation
Modeller can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel. In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes. By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

@@

>>>>

%TODO explain model allocation
%TODO why model allocation
Modeller can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel. In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes. By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

@@

\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parDFS}
\end{figure}

### 2016-06-28 publications/Journal/4b-hotswap.tex
Restructure and write about hotswap component with statistics gathering

Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour might be scarce, might therefore not offer the best possible performance.

\subsection{Extensibility}
%TODO now purely manual
%TODO framework provided to do this using a decider
%TODO possibly using machine learning techniques

\subsection{Statistics Gathering}
%TODO traditionally only access to normal simulation state
%TODO for machine learning might give access to simulator details
%TODO such as ...

\subsubsection{Visualization of Communication}
%TODO explain simple visualizer
%TODO can be used by hotswapper

A user can request a runtime switch between synchronization protocols. A typical use case is when a significant slowdown is observed or memory usage reaches a certain treshold. The simulation engine will swap out the current kernels with new instances while preserving the state of the models and the simulation itself, and resume using the new synchronization protocol. There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on. In a test case with 2 kernels the overhead of a double runtime switch is less than a second.
This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.

%TODO combination with statistics gathering
Dxex optionally collects runtime statistics for each kernel. These record any measurable event in the simulation such as : events created/destroyed, inter/intra kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves. Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure reverted transitions in optimistic synchronization.
The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.

The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or provide feedback to drive synchronization switching.
Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model. These figures demonstrate the effect of allocation visually and show the high inter kernels event count that degrades performance, as we will demonstrate in
\nameref{PholdTreeallocation} . Dotted lines represent connections between models where no events were exchanged, directed edges represents events with frequency as weight, coloring is used to distinguish inter versus intra kernel events.
% Add figure
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
    \label{fig:pholdtree_visualize_seq}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
    \caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
    \label{fig:pholdtree_visualize_parDFS}
\end{figure}

>>>>

Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour might be scarce, might therefore not offer the best possible performance.
In dxex, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
To do this, all kernels are notified of the switch and they are forced to stop simulation using the current synchronization core.
When stopped, each kernel instantiates a new core that is provided with the simulation state of the previous core.
Simulation is then resumed with the new cores after the previous ones are destroyed.

As usual, switching imposes an overhead and should thus only be done if the benefits outweigh the induced overhead.
This overhead depends on the size of the model and the number of simulation cores.
For a simple model and a few cores, the overhead is less than a single second.

Although we currently only support manual switches between different synchronization protocols, this is not necessarily the case.
Ideally, a new component is added to the simulation kernel, which monitors model behaviour and simulation performance, and toggles between them automatically.
Our interface is thus augmented with the necessary bindings for such a component to be defined.
Also, our interface is augmented with an interface for statistics gathering and model behaviour analysis.
The implementation of such a component is currently left open: algorithms can be heavily based on machine learning or similar approaches.

\subsection{Statistics Gathering}
Traditionally, models are not exposed to simulation kernel details due to the wrong level of abstraction.
Simulation models only care about being simulated, and not about how this is being done.
This is different for a new simulation kernel component that has to monitor the behaviour of not only the model, but the simulator as well.

We add performance metrics in the simulation kernel, which logs relevant performance metrics and processes them for use in other components.
These metrics include the number of events created and destroyed, the number of inter and intra kernel events, the number of rollbacks, the measured lookahead, details of the Global Virtual Time (GVT) and Earliest Output Time (EOT) calculations, and basic information on the fairness between different simulation kernels.
With all these metrics, a component can get a fair view on both model and simulation kernel behaviour.

For example, if the actually seen lookahead is significantly higher than the defined lookahead, it might be interesting to switch to optimistic synchronization.
When the number of rollbacks is very high, conservative synchronization might be ideal.
And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation.

For performance reasons, statistics gathering is optional due to the imposed overhead.

\subsubsection{Visualization of Communication}
To provide some more insight in the models we used as benchmarks previously, we created a simple visualization of the simulation trace.
This trace visualizes the allocation of the model and all defined connections.
For each connection, the number of events transferred is annotated.
Examples are shown for the three benchmark models used before: Figures~\ref{fig:Queue_allocation},~\ref{fig:interconnect_allocation_parallel}, and~\ref{fig:phold_allocation} shows traces for the Queue, Interconnect, and PHOLD models respectively.

\begin{figure}
    \center
    \includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
    \caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
    \label{fig:Queue_allocation}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_parallel_allocation.eps}
    \caption{Interconnect parallel simulation trace for 6 models on 3 kernels.}
    \label{fig:interconnect_allocation_parallel}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
    \caption{Phold benchmark trace for parallel simulation using three kernels.}
    \label{fig:phold_allocation}
\end{figure}

Naturally, results similar to this are relevant information that can be used by the hotswapping component.
